---
title: "Blood Pressure"
output:
  html_document:
    toc: true
    theme: united
    df_print: paged
params:
  output_dir: "docs"
---
Blood pressure and pulse readings taken morning and evening with an Omron blood pressure cuff. Also collect binary factors: work, cycling, other exercise and alcohol.

Readings started on Feb 8, 2019.

Read and preprocess blood-pressure file. Join with a file of binary factors.
```{r message=FALSE}
library(lubridate)
```
```{r}
b <- read.csv(file="bp.csv", header=TRUE, sep=",")
b$date <- as.Date(b$date)
bprange <- c(min(b$dia), max(b$sys))
f <- read.csv(file="factors.csv", header=TRUE, sep=",")
f$date <- as.Date(f$date)

# merge blood-pressure and factors on date column
bp <- merge(b, f, by.x = "date", by.y = "date", all.x = TRUE)
bp[is.na(bp)] <- 0
bp$datetime <- as_datetime(paste(bp$date, bp$time))

# plot colours
cdia <- rgb(0,0,1,1/4)
csys <- rgb(1,0,0,1/4)
cpul <- rgb(0,1,0,1/4)
```
A function to plot blood pressure histograms:
```{r}
plot.blood.pressure <- function(bp, title) {
  xmin <- signif(bprange[1] - 5, 1)
  xmax <- signif(bprange[2] + 5, 2)
  b <- seq(xmin, xmax, by=(xmax - xmin) / 20)
  hdia <- hist(bp$dia, breaks=b, plot=F)
  hsys <- hist(bp$sys, breaks=b, plot=F)
  ylim <- c(0, max(c(hsys$counts, hdia$counts)))
  plot(hdia, col=cdia, xlim=c(xmin,xmax), ylim=ylim, main=paste(title, " (", min(bp$date), " to ", max(bp$date), ")", sep=""), xlab="mmHg")
  plot(hsys, col=csys, xlim=c(xmin,xmax), ylim=mc, add=T)
  mdia <- mean(bp$dia)
  msys <- mean(bp$sys)
  abline(v=c(msys, mdia), col=c("red","blue"))
  legend("topright", bty="n", legend=c(paste("sys: ", format(msys, digits=2)), paste("dia: ", format(mdia, digits=2))))
}
```
## Histograms

### Overall
Plot systolic and diastolic pressures for all data points.
```{r}
plot.blood.pressure(bp, "Overall")
```

### Recent
Over the last four weeks:
```{r}
plot.blood.pressure(bp[max(bp$date) - bp$date <= 28,], "Recently")
```

### Morning
```{r}
am <- hour(bp$datetime) < 12
plot.blood.pressure(bp[am,], "Morning")
```

### Evening
```{r}
pm <- hour(bp$datetime) >= 12
plot.blood.pressure(bp[pm,], "Evening")
```

### Work days
```{r}
plot.blood.pressure(bp[bp$work == 1,], "Work")
```

### Days Off
```{r}
plot.blood.pressure(bp[bp$work == 0,], "Days Off")
```

## Other Views

### Over Time
Systolic and diastolic blood pressure over time.
```{r}
movavg <- function(x, n=14) {
  filter(x,rep(1/n,n), sides=2)
}
plot.dates <- function(bp, var, col, yl="mmHg", ma=T) {
  cs <- c("green", "blue")
  dates <- bp$datetime
  points <- bp[,var]
  am <- hour(bp$datetime) < 12
  pm <- !am
  plot(dates[am], points[am], xlab="Date", ylab=yl, main=var, col=cs[1])
  abline(v=dates[bp$work == 1], col="lightgray")
  abline(v=dates[pm & (bp$alcohol == 1)], col="red")
  abline(v=dates[pm & (bp$cycle == 1)], col="green")
  if (ma) {
    lines(dates[am], movavg(points[am]), col=cs[1])
    lines(dates[pm], movavg(points[pm]), col=cs[2])
  }
  sd <- lm(points~dates)
  abline(sd, col=col)
  points(dates[am], points[am], col=cs[1])
  points(dates[pm], points[pm], col=cs[2])
  legend("bottomright", bty="n", legend=paste("r^2: ", format(summary(sd)$r.squared, digits=2)))
  legend("bottomleft", bty="n", legend=c("am", "pm"), text.col = cs, pt.bg=cs)
}
plot.dates(bp, "sys", "red")
plot.dates(bp, "dia", "blue")
```

### Systolic vs Diastolic Pressure

```{r}
plot(bp$dia, bp$sys, xlab = "Diastolic", ylab="Systolic")
msd <- lm(bp$sys ~ bp$dia)
abline(msd)
legend("bottomright", bty="n", legend=paste("r^2: ", format(summary(msd)$r.squared, digits=2)))
```

## Pulse

```{r}
prange <- c(min(bp$pulse), max(bp$pulse))

plot.pulse <- function(bp, title) {
  pmin <- signif(prange[1] - 5, 1)
  pmax <- signif(prange[2] + 5, 2)
  b <- seq(pmin, pmax, by=((pmax - pmin) / 20))
  hpul <- hist(bp$pulse, breaks=b, plot=F)
  plot(hpul, col=cpul, xlim=c(pmin, pmax), main=paste(title, " (", min(bp$date), " to ", max(bp$date), ")", sep=""), xlab="bpm")
  m <- mean(bp$pulse)
  abline(v=m, col=cpul)
  legend("topright", bty="n", legend=paste("av: ", format(m, digits=2)))
}
```
### Overall
```{r}
plot.pulse(bp, "Overall")
```
### Recently
```{r}
plot.pulse(bp[max(bp$date) - bp$date <= 28,], "Recently")
```

### Morning
Working days and days off.
```{r}
plot.pulse(bp[am & bp$work==1 & bp$alcohol==0,], "Morning (work)")
```
```{r}
plot.pulse(bp[am & bp$work==0 & bp$alcohol==0,], "Morning (off)")
```

### Evening
Evenings of working days and days off.
```{r}
plot.pulse(bp[am & bp$work==1 & bp$alcohol==0,], "Evening (work)")
```

```{r}
plot.pulse(bp[pm & bp$work==0 & bp$alcohol==0,], "Evening (off)")
```

### Over Time

```{r}
plot.dates(bp, "pulse", "green", "bpm")
```


## Effects of Alcohol

The effect of alcohol is demonstrated by plotting statistics for several days afterwards.

```{r}
offset.subset <- function(var, val, off) {
  dates <- bp[bp[var] == val, "date"] + off
  return(bp[(bp$date %in% dates),])
}
```

### On Blood Pressure

The zeroth day is the baseline, not much effect expected:
```{r}
plot.blood.pressure(bp[bp$alcohol == 1,], "Day 0")
```

After one day there is a significant increase:
```{r}
plot.blood.pressure(offset.subset("alcohol", 1, 1), "Day 1")
```

On the second day after:
```{r}
plot.blood.pressure(offset.subset("alcohol", 1, 2), "Day 2")
```

On the third day after it plateaus:
```{r}
plot.blood.pressure(offset.subset("alcohol", 1, 3), "Day 3")
```

On the fourth day after it begins to subside:
```{r}
plot.blood.pressure(offset.subset("alcohol", 1, 4), "Day 4")
```

On the fifth day after it has decreased some more:
```{r}
plot.blood.pressure(offset.subset("alcohol", 1, 5), "Day 5")
```

### On Pulse

On the zeroth day pulse is raised:
```{r}
plot.pulse(bp[bp$alcohol == 1,], "Day 0")
```

On the first day it is still raised:
```{r}
plot.pulse(offset.subset("alcohol", 1, 1), "Day 1")
```

By day 2 it has returned to normal:
```{r}
plot.pulse(offset.subset("alcohol", 1, 2), "Day 2")
```

```{r}
plot.pulse(offset.subset("alcohol", 1, 3), "Day 3")
```

### Discussion

The simple model here assumes that alcohol consumed on day 1 will have a residual effect several days later. It does not take into account the effect of consumption on consecutive days.

In fact, alcohol was consumed on consecutive days on a majority of occasions:
```{r}
dates <- f[f$alcohol == 1, "date"]
length(dates[diff(dates) == -1]) / length(dates)
```

The modelling should be updated to take this into account.