---
title: "Blood Pressure"
output:
  html_document:
    toc: true
    theme: united
    df_print: paged
params:
  output_dir: "docs"
---
Blood pressure and pulse readings taken morning and evening with an Omron blood pressure cuff. Some binary factors were also recorded: work, cycling, other exercise and alcohol.

Readings started on Feb 8, 2019.

Read and preprocess blood-pressure file. Join with a file of binary factors.
```{r message=FALSE}
library(lubridate)
```
```{r}
b <- read.csv(file="bp.csv", header=TRUE, sep=",")
b$date <- as.Date(b$date)
bprange <- c(min(b$dia), max(b$sys))
f <- read.csv(file="factors.csv", header=TRUE, sep=",")
f$date <- as.Date(f$date)

# merge blood-pressure and factors on date column
bp <- merge(b, f, by.x = "date", by.y = "date", all.x = TRUE)
bp[is.na(bp)] <- 0
bp$datetime <- as_datetime(paste(bp$date, bp$time))

# plot colours
cdia <- rgb(0,0,1,1/4)
csys <- rgb(1,0,0,1/4)
cpul <- rgb(0,1,0,1/4)
```

A function to plot blood pressure histograms:
```{r}
plot.blood.pressure <- function(bp, title) {
  xmin <- signif(bprange[1] - 5, 1)
  xmax <- signif(bprange[2] + 5, 2)
  b <- seq(xmin, xmax, by=(xmax - xmin) / 20)
  hdia <- hist(bp$dia, breaks=b, plot=F)
  hsys <- hist(bp$sys, breaks=b, plot=F)
  ylim <- c(0, max(c(hsys$counts, hdia$counts)))
  plot(hdia, col=cdia, xlim=c(xmin,xmax), ylim=ylim, main=paste(title, " (", min(bp$date), " to ", max(bp$date), ")", sep=""), xlab="mmHg")
  plot(hsys, col=csys, xlim=c(xmin,xmax), ylim=mc, add=T)
  mdia <- mean(bp$dia)
  msys <- mean(bp$sys)
  abline(v=c(msys, mdia), col=c("red","blue"))
  legend("topright", bty="n", legend=c(paste("sys: ", format(msys, digits=2)), paste("dia: ", format(mdia, digits=2)), paste("n:", nrow(bp))))
}
```
## Histograms

### Overall
Plot systolic and diastolic pressures for all data points.
```{r}
plot.blood.pressure(bp, "Overall")
```

### Recent
Over the last four weeks:
```{r}
plot.blood.pressure(bp[max(bp$date) - bp$date <= 28,], "Recently")
```

### Morning
```{r}
am <- hour(bp$datetime) < 12
plot.blood.pressure(bp[am,], "Morning")
```

### Evening
```{r}
pm <- hour(bp$datetime) >= 12
plot.blood.pressure(bp[pm,], "Evening")
```

### Work days
```{r}
plot.blood.pressure(bp[bp$work == 1,], "Work")
```

### Days Off
```{r}
plot.blood.pressure(bp[bp$work == 0,], "Days Off")
```

## Other Views

### Over Time
Systolic and diastolic blood pressure over time.
```{r}
movavg <- function(x, n=14) {
  filter(x,rep(1/n,n), sides=2)
}
plot.dates <- function(bp, var, col, yl="mmHg", ma=T) {
  cs <- c("green", "blue")
  dates <- bp$datetime
  points <- bp[,var]
  am <- hour(bp$datetime) < 12
  pm <- !am
  plot(dates[am], points[am], xlab="Date", ylab=yl, main=var, col=cs[1])
  abline(v=dates[bp$work == 1], col="lightgray")
  abline(v=dates[pm & (bp$alcohol == 1)], col="red")
  abline(v=dates[pm & (bp$cycle == 1)], col="green")
  if (ma) {
    lines(dates[am], movavg(points[am]), col=cs[1])
    lines(dates[pm], movavg(points[pm]), col=cs[2])
  }
  sd <- lm(points~dates)
  abline(sd, col=col)
  points(dates[am], points[am], col=cs[1])
  points(dates[pm], points[pm], col=cs[2])
  legend("bottomright", bty="n", legend=paste("r^2: ", format(summary(sd)$r.squared, digits=2)))
  legend("bottomleft", bty="n", legend=c("am", "pm"), text.col = cs, pt.bg=cs)
}
plot.dates(bp, "sys", "red")
plot.dates(bp, "dia", "blue")
```

### Systolic vs Diastolic Pressure

```{r}
plot(bp$dia, bp$sys, xlab = "Diastolic", ylab="Systolic")
msd <- lm(bp$sys ~ bp$dia)
abline(msd)
legend("bottomright", bty="n", legend=paste("r^2: ", format(summary(msd)$r.squared, digits=2)))
```

## Pulse

```{r}
prange <- c(min(bp$pulse), max(bp$pulse))

plot.pulse <- function(bp, title) {
  pmin <- signif(prange[1] - 5, 1)
  pmax <- signif(prange[2] + 5, 2)
  b <- seq(pmin, pmax, by=((pmax - pmin) / 20))
  hpul <- hist(bp$pulse, breaks=b, plot=F)
  plot(hpul, col=cpul, xlim=c(pmin, pmax), main=paste(title, " (", min(bp$date), " to ", max(bp$date), ")", sep=""), xlab="bpm")
  m <- mean(bp$pulse)
  abline(v=m, col=cpul)
  legend("topright", bty="n", legend=c(paste("av: ", format(m, digits=2)), paste("n:", nrow(bp))))
}
```

### Overall
```{r}
plot.pulse(bp, "Overall")
```

### Recently
```{r}
plot.pulse(bp[max(bp$date) - bp$date <= 28,], "Recently")
```

### Morning
Working days and days off.
```{r}
plot.pulse(bp[am & bp$work==1 & bp$alcohol==0,], "Morning (work)")
```
```{r}
plot.pulse(bp[am & bp$work==0 & bp$alcohol==0,], "Morning (off)")
```

### Evening
Evenings of working days and days off.
```{r}
plot.pulse(bp[am & bp$work==1 & bp$alcohol==0,], "Evening (work)")
```

```{r}
plot.pulse(bp[pm & bp$work==0 & bp$alcohol==0,], "Evening (off)")
```

### Over Time

```{r}
plot.dates(bp, "pulse", "green", "bpm")
```


## Effects of Alcohol

The effect of alcohol is demonstrated by plotting statistics for several days afterwards.

```{r}
nth.day.after <- function(n) {
  adates <- bp[bp$alcohol == 1, "date"]
  nadates <- bp[(bp$alcohol == 0) & (bp$date %in% (adates + 1)), "date"]
  return(bp[(bp$alcohol == 0) & (bp$date %in% (nadates + n)),])
}
```

### On Blood Pressure

On the first day blood pressure is elevated:
```{r}
plot.blood.pressure(nth.day.after(0), "Day 1")
```

On day 2 it has subsided a bit:
```{r}
plot.blood.pressure(nth.day.after(1), "Day 2")
```

On day 3 it is further reduced:
```{r}
plot.blood.pressure(nth.day.after(2), "Day 3")
```


On the fourth day it is reduced still further:
```{r}
plot.blood.pressure(nth.day.after(3), "Day 4")
```

On the fifth day it has more-or-less returned to normal:
```{r}
plot.blood.pressure(nth.day.after(4), "Day 5")
```

### On Pulse

On the first day pulse is raised:
```{r}
plot.pulse(nth.day.after(0), "Day 1")
```

On the second day it is lowered:
```{r}
plot.pulse(nth.day.after(1), "Day 2")
```

By day 3 it has returned to normal:
```{r}
plot.pulse(nth.day.after(2), "Day 3")
```


### Discussion

The model is complicated by the fact that alcohol tended to be consumed on consecutive days, when it was consumed:
```{r}
dates <- f[f$alcohol == 1, "date"]
length(dates[diff(dates) == -1]) / length(dates)
```

To deal with this, the model considers consecutive days as a single day. It looks for days on which alcohol was consumed, followed by days on which it wasn't. (Such days are "Day 1" in the graphs above.)

The numbers of days per each "day after" illustrate another pattern in the data:
```{r}
ns <- function(n) length(unique(nth.day.after(n)$date))
sapply(c(0, 1, 2, 3, 4, 5, 6), ns, simplify = T)
```

On about half of the occasions when alcohol was consumed, it was also consumed 5 days later (e.g., a Sunday followed by a Friday). On most occasions when it was consumed, it was not consumed during the following 4 days (e.g., Monday to Thursday).

This is confirmed by looking at the days of the week on which it was consumed:
```{r}
table(wday(dates, label=T, abbr=T))
```