---
title: "Time Series Analysis"
output: html_document
---

```{r message=FALSE}
library(pracma)
library(lubridate)
library(tseries)

bp <- read.csv(file="bp.csv", header=TRUE, sep=",")
bp[is.na(bp)] <- 0
bp$datetime <- as_datetime(paste(bp$date, bp$time))
```

We analyse the data in two parts: in the first (WN), the subject was working during the week and abstained from alcohol; in the second (CA), the subject was not working, exeercised during the week, and consumed alcolhol at weekends.

```{r}
wn <- bp[bp$datetime < "2019-04-18",]
ca <- bp[bp$datetime >= "2019-04-18",]
```

### Hurst Exponent (WN)
```{r}
hurstexp(wn$sys)
hurstexp(wn$dia)
hurstexp(wn$pulse)
```

### Augmented Dickey-Fuller Test (WN)

```{r message=FALSE}
adf.test(wn$sys)
adf.test(wn$dia)
adf.test(wn$pulse)
```

All three series are _stationary_.

### ACF and PACF (WN)

```{r}
acf(wn$sys, main="Systolic, WN")
pacf(wn$sys, main="Systolic, WN")
acf(wn$dia, main="Diastolic, WN")
pacf(wn$dia, main="Diastolic, WN")
acf(wn$pulse, main="Pulse, WN")
pacf(wn$pulse, main="Pulse, WN")
```

### Hurst Exponent (CA)
```{r}
hurstexp(ca$sys)
hurstexp(ca$dia)
hurstexp(ca$pulse)
```

All these are solidly over 0.5 indicating the presence of long term memory/ persistence/ auto correlation in the series.  This is what is expected from such a bio time series.

### Augmented Dickey-Fuller Test (CA)

```{r message=FALSE}
adf.test(ca$sys)
adf.test(ca$dia)
adf.test(ca$pulse)
```

All three series are _stationary_.

### ACF and PACF (CA)

```{r}
acf(ca$sys, main="Systolic, CA")
pacf(ca$sys, main="Systolic, CA")
acf(ca$dia, main="Diastolic, CA")
pacf(ca$dia, main="Diastolic, CA")
acf(ca$pulse, main="Pulse, CA")
pacf(ca$pulse, main="Pulse, CA")
```
